Algorithmic detection of conserved quantities of finite-difference schemes for partial differential equations

### 2023-05-02 Papers/Paper ACM/tmp/FCpaper.tex
Friedmann paper

\begin{document}

@@

\section{Introduction}
Many partial differential equations (PDEs) admit integral quantities, that are conserved in time. 
For example, the advection equation preserves energy and the heat equation conserves mass. 
When approximating PDEs by a finite-difference scheme, the question arises whether such quantities are conserved by the scheme. 
Such information can be crucial to estimate whether a scheme approximates a PDE accurately and to determine its stability.\\
While computations for determining conservation are often easy in simple cases, they get rather tedious for more complicated schemes or quantities. 
Hence, automating such computations 
in computer algebra systems is desirable.\\
An algorithm for finding conserved quantities for (continuous) differential equations was developed and implemented  in {\sc Maple} by Cheviakov (see \cite{cheviakov07} and \cite{cheviakov10}).
Hereman et al. (see for example \cite{hereman04} and \cite{hereman08}), proposed an algorithm to compute conserved densities for semi-discretized schemes for PDEs with first-order time derivative. 
Their algorithm, which is implemented in a {\sc Mathematica} package, uses the scaling symmetries of the scheme to construct conserved quantities and calculates their coefficients by using the discrete variational derivative.
Gao et al. extended this algorithm to first-order time-explicit schemes \cite{gao02} and to first-order time-implicit schemes \cite{gao04}. 
They also implemented it in the computer algebra system {\sc Reduce}.\\
In this paper, we propose an algorithm that checks if a quantity is conserved in time under a scheme.
Conserved quantities usually involve integral expressions or their discrete analog, sums.
Gomes et al. proposed algorithms for the simplification of sums in \cite{gomes20}.
They also developed and implemented algorithms for detecting quantities conserved by PDEs and semi-discretized schemes. 
We revise these techniques in Section \ref{assp}. 
To generalize those methods to situations where we do not sum over all arguments of the functions involved, we introduce the discrete partial variational derivative (Section \ref{pvard}).
The main contribution of this paper is an algorithm for checking the conservation of a quantity under a numerical scheme.
Gerdt showed in \cite{gerdt12}, that if the discrete time derivative of the quantity belongs to the difference ideal generated by the scheme, the quantity is conserved. However, some quantities may add to a constant (e.g. telescopic sums) and thus be trivially preserved without belonging to the difference ideal. Moreover, Gerdt's algorithm may not terminate, as the Gr{\"o}bner basis for the difference ideal may not be finite.
We overcome this issues by combining the discrete partial variational derivative with a polynomial ideal (instead of a difference ideal) with finite Gr{\"o}bner basis (Section \ref{algodetec}). 
Our algorithm works for schemes that are explicit and implicit in time and can treat schemes with several and higher-order time  and space derivatives and with several space dimensions.
Additionally, we can handle systems of equations and schemes with parameters. 
We have implemented this algorithm as part of a package in {\sc Mathematica} \cite{wolfram21}.
In the examples, we show that our code finds conserved quantities and proper schemes for the time-implicit and time-explicit discretization of the Burgers equation and a system of PDEs arising in the study of mean-field games (Section \ref{eaa}).

\section{Preliminaries}
\label{assp}
Throughout this section, we follow the ideas and computations in \cite{gomes20}.
Here, subscripts denote indices related to coordinates or tuples while superscripts denote indices related to sequences and families.
To avoid dealing with boundary terms, we work with periodic functions in $\Zz^d$.
\begin{definition}
	Let $N$, $d$ and $m$ be natural numbers. The \emph{discrete torus} is $\Ii:=\{0,\ldots,N-1\}^d\subset \Zz^d$. Define the space
	\begin{gather*}
		\mathcal{P}(\Ii,\Rr^m):=\{u\ |\ u:\Ii\rightarrow\Rr^m\},
	\end{gather*}
	extended to $\Zz^d$ by periodicity.
\end{definition}
\begin{definition}
	The space of functionals $\mathcal{F}(\mathcal{P}(\Ii,\Rr^m))$ (not necessarily linear) on $\mathcal{P}(\Ii,\Rr^m)$ is
\begin{gather*}
	\mathcal{F}(\mathcal{P}(\Ii,\Rr^m)):= \Bigg\{\mathcal{J}:\mathcal{P}(\Ii,\Rr^m)\to \Rr\ \Bigg|\ \mathcal{J}[u]=\sum_{n\in\Ii}F_n[u],\\
	\text{ where }
	F_n[u]=G(u(n+e^1),u(n+e^2),\ldots,u(n+e^k))\\
	\text{ and }\{e^1,\ldots,e^k\}\subset\Zz^d,\ \text{G is smooth}
		\Bigg\}.
\end{gather*}
\end{definition}
In the previous definition, the $e^1,\ldots,e^k$ are not necessarily unit vectors and can vary between functionals. Here, we allow for smooth $G$. However, later we require $G$ to be a polynomial function.
\begin{example}
	\label{functional1}
	Let $d=1$, $m=1$ and define
	\begin{gather*}
		\mathcal{J}[u]:=\sum_{n\in\Ii}(u(n+1)-u(n))^2.
	\end{gather*}
\end{example}
To simplify sums, the discrete variational derivative is a useful tool.
Let $u,v\in\mathcal{P}(\Ii,\Rr^m)$, $\epsilon 
	\in\Rr$ and $\mathcal{J}\in\mathcal{F}(\mathcal{P}(\Ii,\Rr^m))$. Define
	\begin{gather*}
		D\mathcal{J}[u](v):=\frac{d}{d\epsilon}\mathcal{J}[u+\epsilon v]\bigg|_{\epsilon=0}.
	\end{gather*}
	In \eqref{periodicity}, we shift the indices due to periodicity of $u$ and $v$,
	\begin{gather}
	\notag	D\mathcal{J}[u](v)=\frac{d}{d\epsilon}\mathcal{J}[u+\epsilon v]\bigg|_{\epsilon=0}=\frac{d}{d\epsilon}\sum_{n\in\Ii}F_n[u+\epsilon v]\bigg|_{\epsilon=0}\\
	\notag	=\sum_{n\in\Ii}\frac{d}{d\epsilon}G(u(n+e^1)+\epsilon v(n+e^1),\ldots,u(n+e^k)+\epsilon v(n+e^k))\bigg|_{\epsilon=0}\\
	\notag	=\sum_{n\in\Ii}D_1G(u(n+e^1),\ldots,u(n+e^k))v(n+e^1)\\
	\notag	+\ldots+D_kG(u(n+e^1),\ldots,u(n+e^k))v(n+e^k)\\
	\label{periodicity}	=\sum_{n\in\Ii}D_1G(u(n),u(n+e^2-e^1),\ldots,u(n+e^k-e^1))v(n)\\
	\notag	+D_2G(u(n+e^1-e^2),u(n),\ldots,u(n+e^k-e^2))v(n)\\
	\notag	+\ldots+D_kG(u(n+e^1-e^k),u(n+e^2-e^k),\ldots,u(n))v(n)\\
	\notag	=\sum_{n\in\Ii}\Big(D_1G(u(n),u(n+e^2-e^1)\ldots,u(n+e^k-e^1))\\
	\notag	+D_2G(u(n+e^1-e^2),u(n),\ldots,u(n+e^k-e^2))\\
	\notag	+\ldots+D_kG(u(n+e^1-e^k),u(n+e^2-e^k),\ldots,u(n))\Big)v(n)\\
	\notag	=:\sum_{n\in\Ii}\mathcal{V}(\mathcal{J})[u]v(n).	
	\end{gather}
\begin{definition}
$\mathcal{V}(\mathcal{J})$ is the discrete variational derivative of $\mathcal{J}$.
\end{definition}
Because we assume all related functions to be smooth, the discrete variational derivative always exists.
\begin{example}
	Let us return to Example \ref{functional1} and compute
	\begin{gather*}
		D\mathcal{J}[u](v)=\frac{d}{d\epsilon}\mathcal{J}[u+\epsilon v]\bigg|_{\epsilon=0}=\frac{d}{d\epsilon}\sum_{n\in\Ii}F_n[u+\epsilon v]\bigg|_{\epsilon=0}\\
		=\sum_{n\in\Ii}\frac{d}{d\epsilon}G[u(n+1)+\epsilon v(n+1),u(n)+\epsilon v(n)]\bigg|_{\epsilon=0}\\
		=\sum_{n\in\Ii}2(u(n+1)-u(n))v(n+1)+(-2)(u(n+1)-u(n))v(n)\\
		=\sum_{n\in\Ii}2(u(n)-u(n-1))v(n)+(-2)(u(n+1)-u(n))v(n)\\
		=\sum_{n\in\Ii}\Big(2(u(n)-u(n-1))+(-2)(u(n+1)-u(n))\Big)v(n)\\
		=\sum_{n\in\Ii}\Big(-2u(n+1)+4u(n)-2u(n-1)\Big)v(n)\\
		=\sum_{n\in\Ii}\mathcal{V}(\mathcal{J})[u]v(n)
	\end{gather*}
	and hence $\mathcal{V}(\mathcal{J})[u]=-2u(n+1)+4u(n)-2u(n-1)$.
\end{example}

\subsection*{Fundamental theorem}
The algorithms for the simplification of sums presented in \cite{gomes20} rely on the following result:
\begin{teo}\label{theo1}
	Let $\mathcal{J},\tilde{\mathcal{J}}\in \mathcal{F}(\mathcal{P}(\Ii,\Rr^m))$. If $\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[v]=0$ for all $v\in\mathcal{P}(\Ii,\Rr^m)$ and if there exists $u_0\in\mathcal{P}(\Ii,\Rr^m)$ such that $\mathcal{J}[u_0]=\tilde{\mathcal{J}}[u_0]$, then
	\begin{equation*}
		\mathcal{J}[u]=\tilde{\mathcal{J}}[u]
	\end{equation*}
	for all $u\in\mathcal{P}(\Ii,\Rr^m)$.\\
	Conversely,
	\begin{equation*}
		 \mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[u]=0
	\end{equation*}
	for all $u\in\mathcal{P}(\Ii,\Rr^m)$, if $\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$.
\end{teo}
\begin{proof}
	\begin{gather*}
		(\mathcal{J}-\tilde{\mathcal{J}})[u]=(\mathcal{J}-\tilde{\mathcal{J}})[u]-(\mathcal{J}-\tilde{\mathcal{J}})[u_0]\\
		=\int_{0}^{1}\frac{d}{d\lambda}(\mathcal{J}-\tilde{\mathcal{J}})[u_0+\lambda(u-u_0)]\ d\lambda\\
		=\int_{0}^{1}\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[u_0+\lambda(u-u_0)](u-u_0)\ d\lambda=0
	\end{gather*}
	and hence $\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$.
\end{proof}
We use Theorem \ref{theo1} to examine if different sums represent the same quantity.
\begin{example}
	Let $d=m=1$ and consider the functionals
	\begin{gather*}
		\mathcal{J}[u]:=\sum_{n\in\Ii}u(n)u(n+1),\\
		\tilde{\mathcal{J}}[u]:=\sum_{n\in\Ii}u(n-2)u(n-1).
	\end{gather*}
	It is clear, that $\mathcal{J}$ and $\tilde{\mathcal{J}}$ represent the same quantity (by shifting $n$). We confirm this, using the discrete variational derivative:
	\begin{gather*}
		D(\mathcal{J}-\tilde{\mathcal{J}})[u](v)=\\
		=\sum_{n\in\Ii}(u(n+1)+u(n-1)-u(n+1)-u(n-1))v(n)=0.
	\end{gather*}
	Hence, $\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})=0$ and $\mathcal{J}[0]=0=\tilde{\mathcal{J}}[0]$. Therefore, both functionals sum to the same quantity.
\end{example}

@@

	Let $d=2$. It may not be obvious, that

@@

		\sum_{n_1,n_2}u(n_1,n_2-2)^4 - 3 u(n_1,n_2-2)^3 u(n_1+1,n_2-2) \\+ 
		u(n_1,n_2) u(n_1+1,n_2)^3 + 3 u(n_1,n_2-2)^2 u(n_1+1,n_2-2)^2\\ -
		u(n_1,n_2-2) u(n_1+1,n_2-2)^3
		- u(n_1, n_2)^4 +\\
		3 u(n_1,n_2)^3 u(n_1+1,n_2)- 3 u(n_1,n_2)^2 u(n_1+1,n_2)\\
		-3 u(n_1,n_2-2)^2 u(n_1+1,n_2-2)^2+ 3 u(n_1,n_2)^2 u(n_1+1,n_2)=0.

@@

	We confirm this, by computing the discrete variational derivative and noticing that the expression is $0$ for $u_0=0$.

@@

\section{The discrete partial variational derivative}
\label{pvard}
Here, we generalize the discrete variational derivative for situations where we keep one (or several) of the arguments of $u$ constant.
Let $n=(n_1,\ldots,n_d)$ be the variables for the functions in $\mathcal{P}(\Ii,\Rr^m)$. We call $n$ the space variables. Let $l\in\Nn$ and call $t=(t_1,\ldots,t_l)\in\Nn^l$ the time variables. Later in this paper, we only consider the case $l=1$.

@@

	Let $\ \mathcal{P}(\Ii\times\Nn^l,\Rr^m):=\{u\ |\ u(\cdot,t)\in\mathcal{P}(\Ii,\Rr^m)\ \forall t\in\Nn^l \}$, extended to $\Zz^d\times\Nn^l$ by periodicity in the space variables and let $\mathcal{D}(\Nn^l):=\{u\ |\ u:\Nn^l\rightarrow\Rr\}$.

@@

	The space of functions $\mathcal{F}(\mathcal{P}(\Ii\times\Nn^l,\Rr^m),\mathcal{D}(\Nn^l))$ is

@@

		\mathcal{F}(\mathcal{P}(\Ii\times\Nn^l,\Rr^m),\mathcal{D}(\Nn^l)):=\\	
		\Bigg\{\mathcal{J}:\mathcal{P}(\Ii\times\Nn^l)\rightarrow\mathcal{D}(\Nn^l) \Bigg|\ \mathcal{J}[u]=\sum_{n}F_{(n,t)}[u],\\

@@

		\{e^1,\ldots,e^k\}\subset\Zz^{d+l},\text{G is a polynomial in k variables}\Bigg\}.

@@

Here, $(n,t):=(n_1,\ldots,n_d,t_1,\ldots,t_l)\in\Ii\times\Nn^l$.

@@

Throughout this paper, all quantities, whose conservation in time we check, are functions $\mathcal{J}\in\mathcal{F}(\mathcal{P}(\Ii\times\Nn^l,\Rr^m),\mathcal{D}(\Nn^l))$. 
Define $\tilde{n}:=(n_1,...,n_d,s_1,...,s_l)\in\Ii\times\Nn^l$. Using the Kronecker delta $\delta$, we rewrite

@@

	=\sum_{\tilde{n}\in\Ii\times\Nn^l}F_{\tilde{n}}[u]\delta_{(t_1,\ldots,t_{l})}(s_1,\ldots,s_l).

@@

	Let $u,v\in\mathcal{P}(\Ii\times\Nn^l,\Rr^m)$ and compute

@@

	We define $\mathcal{V}^l(\mathcal{J})$ as the discrete partial variational derivative of $\mathcal{J}$.
	Note that $\mathcal{V}^0:=\mathcal{V}$.

@@

	equals $\delta_{t+1}(s)-\delta_t(s)$, because

@@

	Let $u=(u_1,\dots,u_m)\in\mathcal{P}(\Ii\times\Nn,\Rr^m)$. The shift in the $i$-th coordinate for $1\leq i\leq d+1$ by $k\in\Zz$ is

@@

	understanding that $\sigma_{d+1}^k$ shifts the time variable.

@@

	The set of all possible shifts $\Theta$ is

@@

		\Theta:=\{\sigma_1^{k_1}\circ\cdots\circ\sigma_{d+1}^{k_{d+1}}\ |\ k_1,\ldots,k_{d+1}\in\Zz\}.

@@

	Let $\ \mathcal{K}$ be the field generated by $\ \Rr$ and the variables $\{n_1,\ldots,n_d,t\}$. Let $\ \tilde{\mathcal{R}}$ be the polynomial ring over the field $\ \mathcal{K}$ and the variables $\theta\circ u_j(n,t)$ for $\theta\in\Theta$ and $1\leq j\leq m$.

>>>>

\usepackage{color}\begin{document}

@@

\section{Introduction}
Many partial differential equations (PDEs) admit integral quantities, that are conserved in time. 
For example, the advection equation preserves energy and the heat equation conserves mass. 
When approximating PDEs by a finite-difference scheme, the question arises whether such quantities are conserved by the scheme. 
Such information can be crucial to estimate whether a scheme approximates a PDE accurately and to determine its stability.\\
While computations for determining conservation are often easy in simple cases, they get rather tedious for more complicated schemes or quantities. 
Hence, automating such computations 
in computer algebra systems is desirable.\\
An algorithm for finding conserved quantities for (continuous) differential equations was developed and implemented  in {\sc Maple} by Cheviakov (see \cite{cheviakov07} and \cite{cheviakov10}).
Hereman et al. (see for example \cite{hereman04} and \cite{hereman08}), proposed an algorithm to compute conserved densities for semi-discretized schemes for PDEs with first-order time derivative. 
Their algorithm, which is implemented in a {\sc Mathematica} package, uses the scaling symmetries of the scheme to construct conserved quantities and calculates their coefficients by using the discrete variational derivative.
Gao et al. extended this algorithm to first-order time-explicit schemes \cite{gao02} and to first-order time-implicit schemes \cite{gao04}. 
They also implemented it in the computer algebra system {\sc Reduce}.\\
In this paper, we propose an algorithm that checks if a quantity is conserved in time under a scheme.
Conserved quantities usually involve integral expressions or their discrete analog, sums.
Gomes et al. proposed algorithms for the simplification of sums in \cite{gomes20}.
They also developed and implemented algorithms for detecting quantities conserved by PDEs and semi-discretized schemes. 
We revise these techniques in Section \ref{assp}
 {\color{red} Forward or undefined reference: \begin{verbatim}\ref{assp}\end{verbatim}.}
. 
To generalize those methods to situations where we do not sum over all arguments of the functions involved, we introduce the discrete partial variational derivative (Section \ref{pvard}
 {\color{red} Forward or undefined reference: \begin{verbatim}\ref{pvard}\end{verbatim}.}
).
The main contribution of this paper is an algorithm for checking the conservation of a quantity under a numerical scheme.
Gerdt showed in \cite{gerdt12}, that if the discrete time derivative of the quantity belongs to the difference ideal generated by the scheme, the quantity is conserved. However, some quantities may add to a constant (e.g. telescopic sums) and thus be trivially preserved without belonging to the difference ideal. Moreover, Gerdt's algorithm may not terminate, as the Gr{\"o}bner basis for the difference ideal may not be finite.
We overcome this issues by combining the discrete partial variational derivative with a polynomial ideal (instead of a difference ideal) with finite Gr{\"o}bner basis (Section \ref{algodetec}
 {\color{red} Forward or undefined reference: \begin{verbatim}\ref{algodetec}\end{verbatim}.}
). 
Our algorithm works for schemes that are explicit and implicit in time and can treat schemes with several and higher-order time  and space derivatives and with several space dimensions.
Additionally, we can handle systems of equations and schemes with parameters. 
We have implemented this algorithm as part of a package in {\sc Mathematica} \cite{wolfram21}.
In the examples, we show that our code finds conserved quantities and proper schemes for the time-implicit and time-explicit discretization of the Burgers equation and a system of PDEs arising in the study of mean-field games (Section \ref{eaa}
 {\color{red} Forward or undefined reference: \begin{verbatim}\ref{eaa}\end{verbatim}.}
).

\section{Preliminaries}
\label{assp}
Throughout this section, we follow the ideas and computations in \cite{gomes20}.
Here, subscripts denote indices related to coordinates or tuples while superscripts denote indices related to sequences and families.
To avoid dealing with boundary terms, we work with periodic functions in \({\mathbb{Z}}^d\)
{\color{blue} New variable(s): \(d\), \(Z\),  }
.
	Let \(N\)
{\color{blue} New variable(s): \(N\),  }
, \(d\) and \(m\)
{\color{blue} New variable(s): \(m\),  }
 be natural numbers. The \emph{discrete torus} is \({\mathcal{I}}:=\{0,\ldots,N-1\}^d\subset {\mathbb{Z}}^d\). Define the space
		\mathcal{P}({\mathcal{I}},{\mathbb{R}}^m):=\{u\ |\ u:{\mathcal{I}}\rightarrow{\mathbb{R}}^m\},
	extended to \({\mathbb{Z}}^d\) by periodicity.
	The space of functionals \(\mathcal{F}(\mathcal{P}({\mathcal{I}},{\mathbb{R}}^m))\)
{\color{blue} New variable(s): \(F\), \(P\), \(R\),  }
 (not necessarily linear) on \(\mathcal{P}({\mathcal{I}},{\mathbb{R}}^m)\) is
	\mathcal{F}(\mathcal{P}({\mathcal{I}},{\mathbb{R}}^m)):= \Bigg\{\mathcal{J}:\mathcal{P}({\mathcal{I}},{\mathbb{R}}^m)\to {\mathbb{R}}\ \Bigg|\ \mathcal{J}[u]=\sum_{n\in{\mathcal{I}}}F_n[u],\\
	\text{ where }
	F_n[u]=G(u(n+e^1),u(n+e^2),\ldots,u(n+e^k))\\
	\text{ and }\{e^1,\ldots,e^k\}\subset{\mathbb{Z}}^d,\ \text{G is smooth}
In the previous definition, the \(e^1,\ldots,e^k\)
{\color{blue} New variable(s): \(k\),  }
 are not necessarily unit vectors and can vary between functionals. Here, we allow for smooth \(G\)
{\color{blue} New variable(s): \(G\),  }
. However, later we require \(G\) to be a polynomial function.

@@

	\label{functional1}
	Let \(d=1\), \(m=1\) and define

@@

		\mathcal{J}[u]:=\sum_{n\in{\mathcal{I}}}(u(n+1)-u(n))^2.

@@

@@

To simplify sums, the discrete variational derivative is a useful tool.
Let \(u,v\in\mathcal{P}({\mathcal{I}},{\mathbb{R}}^m)\)
{\color{blue} New variable(s): \(u\), \(v\),  }
, \(\epsilon 
	\in{\mathbb{R}}\)
{\color{blue} New variable(s): \(\epsilon\),  }
 and \(\mathcal{J}\in\mathcal{F}(\mathcal{P}({\mathcal{I}},{\mathbb{R}}^m))\)
{\color{blue} New variable(s): \(J\),  }
. Define
		D\mathcal{J}[u](v):=\frac{d}{d\epsilon}\mathcal{J}[u+\epsilon v]\bigg|_{\epsilon=0}.
	In \eqref{periodicity}
 {\color{red} Forward or undefined reference: \begin{verbatim}\eqref{periodicity}\end{verbatim}.}
, we shift the indices due to periodicity of \(u\) and \(v\),
	\begin{gather}
	\notag	D\mathcal{J}[u](v)=\frac{d}{d\epsilon}\mathcal{J}[u+\epsilon v]\bigg|_{\epsilon=0}=\frac{d}{d\epsilon}\sum_{n\in{\mathcal{I}}}F_n[u+\epsilon v]\bigg|_{\epsilon=0}\\
	\notag	=\sum_{n\in{\mathcal{I}}}\frac{d}{d\epsilon}G(u(n+e^1)+\epsilon v(n+e^1),\ldots,u(n+e^k)+\epsilon v(n+e^k))\bigg|_{\epsilon=0}\\
	\notag	=\sum_{n\in{\mathcal{I}}}D_1G(u(n+e^1),\ldots,u(n+e^k))v(n+e^1)\\
	\notag	+\ldots+D_kG(u(n+e^1),\ldots,u(n+e^k))v(n+e^k)\\
	\label{periodicity}	=\sum_{n\in{\mathcal{I}}}D_1G(u(n),u(n+e^2-e^1),\ldots,u(n+e^k-e^1))v(n)\\
	\notag	+D_2G(u(n+e^1-e^2),u(n),\ldots,u(n+e^k-e^2))v(n)\\
	\notag	+\ldots+D_kG(u(n+e^1-e^k),u(n+e^2-e^k),\ldots,u(n))v(n)\\
	\notag	=\sum_{n\in{\mathcal{I}}}\Big(D_1G(u(n),u(n+e^2-e^1)\ldots,u(n+e^k-e^1))\\
	\notag	+D_2G(u(n+e^1-e^2),u(n),\ldots,u(n+e^k-e^2))\\
	\notag	+\ldots+D_kG(u(n+e^1-e^k),u(n+e^2-e^k),\ldots,u(n))\Big)v(n)\\
	\notag	=:\sum_{n\in{\mathcal{I}}}\mathcal{V}(\mathcal{J})[u]v(n).	
	\end{gather}

@@

\(\mathcal{V}(\mathcal{J})\)
{\color{blue} New variable(s): \(V\),  }
 is the discrete variational derivative of \(\mathcal{J}\).
Because we assume all related functions to be smooth, the discrete variational derivative always exists.
	Let us return to Example \ref{functional1} and compute
		D\mathcal{J}[u](v)=\frac{d}{d\epsilon}\mathcal{J}[u+\epsilon v]\bigg|_{\epsilon=0}=\frac{d}{d\epsilon}\sum_{n\in{\mathcal{I}}}F_n[u+\epsilon v]\bigg|_{\epsilon=0}\\
		=\sum_{n\in{\mathcal{I}}}\frac{d}{d\epsilon}G[u(n+1)+\epsilon v(n+1),u(n)+\epsilon v(n)]\bigg|_{\epsilon=0}\\
		=\sum_{n\in{\mathcal{I}}}2(u(n+1)-u(n))v(n+1)+(-2)(u(n+1)-u(n))v(n)\\
		=\sum_{n\in{\mathcal{I}}}2(u(n)-u(n-1))v(n)+(-2)(u(n+1)-u(n))v(n)\\
		=\sum_{n\in{\mathcal{I}}}\Big(2(u(n)-u(n-1))+(-2)(u(n+1)-u(n))\Big)v(n)\\
		=\sum_{n\in{\mathcal{I}}}\Big(-2u(n+1)+4u(n)-2u(n-1)\Big)v(n)\\
		=\sum_{n\in{\mathcal{I}}}\mathcal{V}(\mathcal{J})[u]v(n)
	and hence \(\mathcal{V}(\mathcal{J})[u]=-2u(n+1)+4u(n)-2u(n-1)\).

\subsection*{Fundamental theorem}
The algorithms for the simplification of sums presented in \cite{gomes20} rely on the following result:
\begin{teo}\label{theo1}
	Let \(\mathcal{J},\tilde{\mathcal{J}}\in \mathcal{F}(\mathcal{P}({\mathcal{I}},{\mathbb{R}}^m))\). If \(\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[v]=0\) for all \(v\in\mathcal{P}({\mathcal{I}},{\mathbb{R}}^m)\) and if there exists \(u_0\in\mathcal{P}({\mathcal{I}},{\mathbb{R}}^m)\)
{\color{blue} New variable(s): \(u_0\),  }
 such that \(\mathcal{J}[u_0]=\tilde{\mathcal{J}}[u_0]\), then
	\begin{equation*}
		\mathcal{J}[u]=\tilde{\mathcal{J}}[u]
	\end{equation*}
	for all \(u\in\mathcal{P}({\mathcal{I}},{\mathbb{R}}^m)\).\\
	Conversely,
	\begin{equation*}
		 \mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[u]=0
	\end{equation*}
	for all \(u\in\mathcal{P}({\mathcal{I}},{\mathbb{R}}^m)\), if \(\mathcal{J}[u]=\tilde{\mathcal{J}}[u]\).
\end{teo}
\begin{proof}
	\begin{gather*}
		(\mathcal{J}-\tilde{\mathcal{J}})[u]=(\mathcal{J}-\tilde{\mathcal{J}})[u]-(\mathcal{J}-\tilde{\mathcal{J}})[u_0]\\
		=\int_{0}^{1}\frac{d}{d\lambda}(\mathcal{J}-\tilde{\mathcal{J}})[u_0+\lambda(u-u_0)]\ d\lambda\\
		=\int_{0}^{1}\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[u_0+\lambda(u-u_0)](u-u_0)\ d\lambda=0
	\end{gather*}
	and hence \(\mathcal{J}[u]=\tilde{\mathcal{J}}[u]\).
\end{proof}
We use Theorem \ref{theo1} to examine if different sums represent the same quantity.
\begin{example}
	Let \(d=m=1\) and consider the functionals
	\begin{gather*}
		\mathcal{J}[u]:=\sum_{n\in{\mathcal{I}}}u(n)u(n+1),\\
		\tilde{\mathcal{J}}[u]:=\sum_{n\in{\mathcal{I}}}u(n-2)u(n-1).
	\end{gather*}
	It is clear, that \(\mathcal{J}\) and \(\tilde{\mathcal{J}}\) represent the same quantity (by shifting \(n\)
{\color{blue} New variable(s): \(n\),  }
). We confirm this, using the discrete variational derivative:
	\begin{gather*}
		D(\mathcal{J}-\tilde{\mathcal{J}})[u](v)=\\
		=\sum_{n\in{\mathcal{I}}}(u(n+1)+u(n-1)-u(n+1)-u(n-1))v(n)=0.
	\end{gather*}
	Hence, \(\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})=0\) and \(\mathcal{J}[0]=0=\tilde{\mathcal{J}}[0]\). Therefore, both functionals sum to the same quantity.
\end{example}

\begin{example}
	Let \(d=2\). It may not be obvious, that
	\begin{gather*}
		\sum_{n_1,n_2}u(n_1,n_2-2)^4 - 3 u(n_1,n_2-2)^3 u(n_1+1,n_2-2) \\+ 
		u(n_1,n_2) u(n_1+1,n_2)^3 + 3 u(n_1,n_2-2)^2 u(n_1+1,n_2-2)^2\\ -
		u(n_1,n_2-2) u(n_1+1,n_2-2)^3
		- u(n_1, n_2)^4 +\\
		3 u(n_1,n_2)^3 u(n_1+1,n_2)- 3 u(n_1,n_2)^2 u(n_1+1,n_2)\\
		-3 u(n_1,n_2-2)^2 u(n_1+1,n_2-2)^2+ 3 u(n_1,n_2)^2 u(n_1+1,n_2)=0.
	\end{gather*}
	We confirm this, by computing the discrete variational derivative and noticing that the expression is \(0\)
{\color{blue} New variable(s): \(\text{Null}\),  }
 for \(u_0=0\).
\end{example}
\section{The discrete partial variational derivative}
\label{pvard}
Here, we generalize the discrete variational derivative for situations where we keep one (or several) of the arguments of \(u\) constant.
Let \(n=(n_1,\ldots,n_d)\)
{\color{blue} New variable(s): \(n_1\), \(n_d\),  }
 be the variables for the functions in \(\mathcal{P}({\mathcal{I}},{\mathbb{R}}^m)\). We call \(n\) the space variables. Let \(l\in{\mathbb{N}}\)
{\color{blue} New variable(s): \(l\),  }
 and call \(t=(t_1,\ldots,t_l)\in{\mathbb{N}}^l\)
{\color{blue} New variable(s): \(t\), \(t_1\), \(t_l\),  }
 the time variables. Later in this paper, we only consider the case \(l=1\).
\begin{definition}
	Let {\color{red} \(\ \mathcal{P}({\mathcal{I}}\times{\mathbb{N}}^l,{\mathbb{R}}^m):=\{u\ |\ u(\cdot,t)\in\mathcal{P}({\mathcal{I}},{\mathbb{R}}^m)\ \forall t\in{\mathbb{N}}^l \}\)

  Equation errors and warnings:

 Unbalanced $|\hdots|$}

, extended to \({\mathbb{Z}}^d\times{\mathbb{N}}^l\) by periodicity in the space variables and let {\color{red} \(\mathcal{D}({\mathbb{N}}^l):=\{u\ |\ u:{\mathbb{N}}^l\rightarrow{\mathbb{R}}\}\)

  Equation errors and warnings:

 Unbalanced $|\hdots|$}

{\color{blue} New variable(s): \(D\),  }
.

@@

	The space of functions \(\mathcal{F}(\mathcal{P}({\mathcal{I}}\times{\mathbb{N}}^l,{\mathbb{R}}^m),\mathcal{D}({\mathbb{N}}^l))\) is

@@

		\mathcal{F}(\mathcal{P}({\mathcal{I}}\times{\mathbb{N}}^l,{\mathbb{R}}^m),\mathcal{D}({\mathbb{N}}^l)):=\\	
		\Bigg\{\mathcal{J}:\mathcal{P}({\mathcal{I}}\times{\mathbb{N}}^l)\rightarrow\mathcal{D}({\mathbb{N}}^l) \Bigg|\ \mathcal{J}[u]=\sum_{n}F_{(n,t)}[u],\\

@@

		\{e^1,\ldots,e^k\}\subset{\mathbb{Z}}^{d+l},\text{G is a polynomial in k variables}\Bigg\}.

@@

Here, \((n,t):=(n_1,\ldots,n_d,t_1,\ldots,t_l)\in{\mathcal{I}}\times{\mathbb{N}}^l\).

@@

Throughout this paper, all quantities, whose conservation in time we check, are functions \(\mathcal{J}\in\mathcal{F}(\mathcal{P}({\mathcal{I}}\times{\mathbb{N}}^l,{\mathbb{R}}^m),\mathcal{D}({\mathbb{N}}^l))\). 
Define \(\tilde{n}:=(n_1,...,n_d,s_1,...,s_l)\in{\mathcal{I}}\times{\mathbb{N}}^l\)
{\color{blue} New variable(s): \(s_1\), \(s_l\),  }
. Using the Kronecker delta \(\delta\)
{\color{blue} New variable(s): \(\delta\),  }
, we rewrite

@@

	=\sum_{\tilde{n}\in{\mathcal{I}}\times{\mathbb{N}}^l}F_{\tilde{n}}[u]\delta_{(t_1,\ldots,t_{l})}(s_1,\ldots,s_l).

@@

	Let \(u,v\in\mathcal{P}({\mathcal{I}}\times{\mathbb{N}}^l,{\mathbb{R}}^m)\) and compute

@@

	We define \(\mathcal{V}^l(\mathcal{J})\) as the discrete partial variational derivative of \(\mathcal{J}\).
	Note that \(\mathcal{V}^0:=\mathcal{V}\).

@@

	equals \(\delta_{t+1}(s)-\delta_t(s)\)
{\color{blue} New variable(s): \(s\), \(\delta _t\),  }
, because

@@

	Let \(u=(u_1,\dots,u_m)\in\mathcal{P}({\mathcal{I}}\times{\mathbb{N}},{\mathbb{R}}^m)\)
{\color{blue} New variable(s): \(u_1\), \(u_m\),  }
. The shift in the \(i\)-th coordinate for \(1\leq i\leq d+1\) by \(k\in{\mathbb{Z}}\) is

@@

	understanding that \(\sigma_{d+1}^k\)
{\color{blue} New variable(s): \(\sigma\),  }
 shifts the time variable.

@@

	The set of all possible shifts \(\Theta\)
{\color{blue} New variable(s): \(\Theta\),  }
 is

@@

		\Theta:=\{\sigma_1^{k_1}\circ\cdots\circ\sigma_{d+1}^{k_{d+1}}\ |\ k_1,\ldots,k_{d+1}\in{\mathbb{Z}}\}.

@@

	Let \(\ \mathcal{K}\)
{\color{blue} New variable(s): \(K\),  }
 be the field generated by \(\ {\mathbb{R}}\) and the variables \(\{n_1,\ldots,n_d,t\}\). Let \(\ \tilde{\mathcal{R}}\) be the polynomial ring over the field \(\ \mathcal{K}\) and the variables \(\theta\circ u_j(n,t)\)
{\color{blue} New variable(s): \(\theta\),  }
 for \(\theta\in\Theta\) and \(1\leq j\leq m\).
